JuPOETs: A Constrained Multiobjective Optimization Approach to Estimate Biochemical Model Ensembles in the Julia Programming Language

### 2016-05-29 Paper_BioArXiv_Submitted_v1.tex
Updated the BMC version

JuPOETs integrates simulated annealing with Pareto optimality to estimate parameter sets on or near the optimal tradeoff surface between competing training objectives
(Fig. \ref{fig:fig-MOSAalgorithm} and Algorithm \ref{algo:JuPOETs-pcode}).
The central idea of POETs is a mapping between the value of the objective vector evaluated at $\mathbf{p}_{i+1}$ (parameter guess at iteration $i+1$) and Pareto rank.
JuPOETs calculates the performance of a candidate parameter set $\mathbf{p}_{i+1}$ by calling the user defined \texttt{objective} function; \texttt{objective} takes
a parameter set as an input and returns a $\mathcal{K}\times{1}$ objective vector. Candidate parameter sets are generated by the user supplied \texttt{neighbor} function.
The error vector associated with $\mathbf{p}_{i+1}$ is ranked using the builtin Pareto \texttt{rank} function, by comparing the current error at iteration $i$+1 to the error archive $\mathcal{O}_{i}$ (all error vectors up to iteration $i$-1 that meeting a ranking criteria). Pareto rank is a measure of distance from the trade-off surface;
parameter sets on or near the optimal trade-off surface between the objectives have a rank equal to $0$ (no other current parameter sets are better).
Sets with increasing non-zero rank are progressively further away from the optimal trade-off surface.
Thus, a parameter set with a rank $=0$ is \emph{better} in a trade-off sense than rank $>0$.
We implemented the Fonseca and Fleming ranking scheme \cite{RANKING} in the builtin \texttt{rank} function:
\begin{equation}\label{eqn_rank}
\texttt{rank}\left(\mathbf{p}_{i+1}\mid \mathcal{O}_{i}\right)=r
\end{equation} where rank $r$ is the number of parameter sets that dominate (are better than) parameter set $\mathbf{p}_{i+1}$.
We used the Pareto rank to inform the SA calculation.
The parameter set $\mathbf{p}_{i+1}$ was accepted or rejected by the SA, by calculating an \texttt{acceptance} probability $\mathcal{P}\left(\mathbf{p}_{i+1}\right)$:
\begin{equation}\label{eqn_costMOSA}
\mathcal{P}(\mathbf{p}_{i+1}) \equiv \exp{\{-\texttt{rank}\left(\mathbf{p}_{i+1} \mid \mathcal{S}_{i} \right)/T\}}
\end{equation}
where $T$ is the computational annealing temperature.
As $\texttt{rank}\left(\mathbf{p}_{i+1}\mid \mathcal{O}_{i}\right)\rightarrow{0}$, the acceptance probability moves toward one,
ensuring that we explore parameter sets along the Pareto surface.
Occasionally, (depending upon $T$) a parameter set with a high Pareto rank was accepted by the SA allowing a more diverse search of the parameter space.
However, as $T$ is reduced, the probability of accepting a high-rank set occurring decreases.
Parameter sets could also be accepted by the SA but $\emph{not}$ permanently archived in $\mathcal{S}_{i}$.
Only parameter sets with rank less than or equal to threshold (rank $\leq$4 by default) were included in $\mathcal{S}_{i}$, where the archive was re-ranked and filtered after
every new parameter set was accepted.
Parameter bounds were implemented in the \texttt{neighbor} function as box constraints, while problem specific constraints were implemented in \texttt{objective} using a penalty method:
\begin{equation}
  O_{i}+\lambda\sum_{j=1}^{\mathcal{C}}\min\Bigl\{0,g_j\left(t,\mathbf{x}(t,\mathbf{p}),\mathbf{u}(t),\mathbf{p}\right)\Bigr\}\qquad~i=1,\hdots,\mathcal{K}
\end{equation}where $\lambda$ denotes the penalty parameter ($\lambda$ = 100 by default).
However, because both the neighbor and objective functions are user defined, different constraint implementations are easily defined.
JuPOETs can be installed using the Julia package manager from the JuPOETs repository at https://github.com/varnerlab/POETs.jl.
Sample code is included in the \texttt{sample/biochemical} subdirectory of the JuPOETs repository to help users get started using JuPOETs in their projects.

@@

\begin{figure}[h]
  \includegraphics[width=1.0\textwidth]{./figs/Figure-4-BiochemicalProofOfConcept.pdf}
  \caption{Proof of concept biochemical network study.
  Inset right: Prototypical biochemical network with six metabolites and seven reactions modeled using the hybrid cybernetic approach (HCM).
  Intracellular cellmass precursors $A,B$, and $C$ are balanced (no accumulation) while the extracellular metabolites $A_{e},B_{e}$, and $C_{e}$ are dynamic.
  The oval denotes the cell boundary, $q_{j}$ is the $j$th flux across the boundary, and $v_{k}$ denotes the $k$th intracellular flux.
  Four data sets (each with $A_{e},B_{e}$,$C_{e}$ and cellmass measurements) were generated by varying the kinetic constants for each biochemical mode.
  Each data set was a single objective in the JuPOETs procedure.
  A: Ensemble simulation of extracellular substrate $A_{e}$ and cellmass versus time.
  B: Ensemble simulation of extracellular substrate $B_{e}$ and $C_{e}$ versus time.
  The gray region denotes the 95\% confidence estimate of the mean ensemble simulation.
  The data points denote mean synthetic measurements, while the error bars denote the 95\% confidence estimate of the measurement computed over the four training data sets.
  C: Trade-off plots between the four training objectives. The quantity $O_{j}$ denotes the jth training objective.
  Each point represents a member of the parameter ensemble, where gray denotes rank 0 sets, while black denotes rank 1 sets.
   }\label{fig:fig-biochemical}
\end{figure}

@@

>>>>

JuPOETs integrates simulated annealing with Pareto optimality to estimate parameter sets on or near the optimal tradeoff surface between competing training objectives
(Fig. \ref{fig:fig-MOSAalgorithm} and Algorithm \ref{algo:JuPOETs-pcode}).
The central idea of POETs is a mapping between the value of the objective vector evaluated at $\mathbf{p}_{i+1}$ (parameter guess at iteration $i+1$) and Pareto rank.
JuPOETs calculates the performance of a candidate parameter set $\mathbf{p}_{i+1}$ by calling the user defined \texttt{objective} function; \texttt{objective} takes
a parameter set as an input and returns a $\mathcal{K}\times{1}$ objective vector. Candidate parameter sets are generated by the user supplied \texttt{neighbor} function.
The error vector associated with $\mathbf{p}_{i+1}$ is ranked using the builtin Pareto \texttt{rank} function, by comparing the current error at iteration $i+1$ to the error archive $\mathcal{O}_{i}$ (all error vectors up to iteration $i-1$ meeting a ranking criteria). Pareto rank is a measure of distance from the trade-off surface;
parameter sets on or near the optimal trade-off surface between the objectives have a rank equal to $0$ (no other current parameter sets are better).
Sets with increasing non-zero rank are progressively further away from the optimal trade-off surface.
Thus, a parameter set with a rank $=0$ is \emph{better} in a trade-off sense than rank $>0$.
We implemented the Fonseca and Fleming ranking scheme in the builtin \texttt{rank} function \cite{RANKING}:
\begin{equation}\label{eqn_rank}
\texttt{rank}\left(\mathcal{O}_{i+1}\left(\mathbf{p}_{i+1}\right)\mid \mathcal{O}_{i}\right)=r
\end{equation} where rank $r$ is the number of parameter sets that dominate (are better than) parameter set $\mathbf{p}_{i+1}$, and $\mathcal{O}_{i+1}\left(\mathbf{p}_{i+1}\right)$
denotes the objective vector evaluated at $\mathbf{p}_{i+1}$.
We used the Pareto rank to inform the SA calculation.
The parameter set $\mathbf{p}_{i+1}$ was accepted or rejected by the SA, by calculating an \texttt{acceptance} probability $\mathcal{P}\left(\mathbf{p}_{i+1}\right)$:
\begin{equation}\label{eqn_costMOSA}
\mathcal{P}(\mathbf{p}_{i+1}) \equiv \exp{\{-\texttt{rank}\left(\mathcal{O}_{i+1}\left(\mathbf{p}_{i+1}\right) \mid \mathcal{O}_{i} \right)/T\}}
\end{equation}
where $T$ is the computational annealing temperature.
As $\texttt{rank}\left(\mathcal{O}_{i+1}\left(\mathbf{p}_{i+1}\right)\mid \mathcal{O}_{i}\right)\rightarrow{0}$, the acceptance probability moves toward one,
ensuring that we explore parameter sets along the Pareto surface.
Occasionally, (depending upon $T$) a parameter set with a high Pareto rank was accepted by the SA allowing a more diverse search of the parameter space.
However, as $T$ is reduced, the probability of accepting a high-rank set occurring decreases.
Parameter sets could also be accepted by the SA but $\emph{not}$ permanently archived in $\mathcal{S}_{i}$.
Only parameter sets with rank less than or equal to threshold (rank $\leq$4 by default) were included in $\mathcal{S}_{i}$, where the archive was re-ranked and filtered after
every new parameter set was accepted.
Parameter bounds were implemented in the \texttt{neighbor} function as box constraints, while problem specific constraints were implemented in \texttt{objective} using a penalty method:
\begin{equation}
  O_{i}+\lambda\sum_{j=1}^{\mathcal{C}}\min\Bigl\{0,g_j\left(t,\mathbf{x}(t,\mathbf{p}),\mathbf{u}(t),\mathbf{p}\right)\Bigr\}\qquad~i=1,\hdots,\mathcal{K}
\end{equation}where $\lambda$ denotes the penalty parameter ($\lambda$ = 100 by default).
However, because both the neighbor and objective functions are user defined, different constraint implementations are easily defined.
JuPOETs can be installed using the Julia package manager from the JuPOETs repository at https://github.com/varnerlab/POETs.jl.
Sample code is included in the \texttt{sample/biochemical} subdirectory of the JuPOETs repository to help users get started using JuPOETs in their projects.

@@

\begin{figure}[h]
  \includegraphics[width=1.0\textwidth]{./figs/Figure-4-BiochemicalProofOfConcept.pdf}
  \caption{Proof of concept biochemical network study.
  Inset right: Prototypical biochemical network with six metabolites and seven reactions modeled using the hybrid cybernetic approach (HCM).
  Intracellular cellmass precursors $A,B$, and $C$ are balanced (no accumulation) while the extracellular metabolites $A_{e},B_{e}$, and $C_{e}$ are dynamic.
  The oval denotes the cell boundary, $q_{j}$ is the $j$th flux across the boundary, and $v_{k}$ denotes the $k$th intracellular flux.
  Four data sets (each with $A_{e},B_{e}$,$C_{e}$ and cellmass measurements) were generated by varying the kinetic constants for each biochemical mode.
  Each data set was a single objective in the JuPOETs procedure.
  A: Ensemble simulation of extracellular substrate $A_{e}$ and cellmass versus time.
  B: Ensemble simulation of extracellular substrate $B_{e}$ and $C_{e}$ versus time.
  The gray region denotes the 95\% confidence estimate of the mean ensemble simulation.
  The data points denote mean synthetic measurements, while the error bars denote the 95\% confidence estimate of the measurement computed over the four training data sets.
  C: Trade-off plots between the four training objectives. The quantity $O_{j}$ denotes the jth training objective.
  Each point represents a member of the parameter ensemble, where black denotes a rank 0 set, while gray denotes rank 1 set.
   }\label{fig:fig-biochemical}
\end{figure}

@@

\begin{table}
	\includegraphics[width=1.0\textwidth]{./figs/Table-1-Objectives.pdf}
	\caption{Multi-objective optimization test problems. We tested the JuPOETs implementation on three two-dimensional test problems, with one-, two- and three-dimensional parameter vectors.
  Each problem had parameter bounds constraints, however, on the Binh and Korn function had additional non-linear problem constraints. For the Fonesca and Fleming problem, N = 3.}
  \label{table:objective-table}
\end{table}

### 2016-05-29 Paper_v1.tex
Updated the BMC version

\begin{abstract} % abstract
Ensemble modeling is a well established approach for obtaining robust predictions and course grained population behavior in large deterministic mathematical models.
In this study, we present the Pareto Optimal Ensemble Technique in the Julia programming language (JuPOETs).
JuPOETs integrates simulated annealing with Pareto optimality to estimate an ensemble of parameter sets on or near the optimal tradeoff surface between competing training objectives.
We demonstrated JuPOETs on a suite of multiobjective problems, including test functions with parameter bounds and system constraints as well as identification of a proof-of-concept biochemical model.
JuPOETs identified optimal or near optimal solutions approximately six-fold faster than a corresponding implementation in Octave.
JuPOETs can be installed using the Julia package manager from the JuPOETs GitHub repository at https://github.com/varnerlab/POETs.jl.
\end{abstract}

@@

\section*{Conclusions}

JuPOETs is a significant advance over the previous POETs implementation.
It offers improved performance and is highly adaptable to different problem types.
We demonstrated JuPOETs on a suite of test problems, and a proof-of-concept biochemical model.
However, there are several areas that could be explored further to improve JuPOETs.
First, JuPOETs should be compared with other multiobjective evolutionary algorithms (MOEAs) to determine its relative performance on test and real world problems.
Many evolutionary approaches e.g., the nondominated sorting genetic algorithm (NSGA) family of algorithms, have been adapted to solve multiobjective optimization problems \cite{DEB2002,HUBAND2006}. It is unclear if JuPOETs will perform as well as these other approaches;
one potential advantage that JuPOETs may have is the local refinement step which temporarily reduces the problem to a single objective formulation.
Previously, this hybrid approach led to better convergence on a proof-of-concept signal transduction model \cite{Song:2010aa}.
For many real world parameter estimation problems, the bulk of the execution time is spent evaluating the objective functions.
One strategy to improve performance could be to optimize surrogates \cite{SURROGATES}, while another would be parallel execution of the objective functions.
Currently, JuPOETs serially evaluates the objective function vector.
However, parallel evaluation of the objective functions could be easily implemented using a variety of techniques without changing the main run loop of JuPOETs.
Because of the flexible function pointer architecture of JuPOETs, the only changes required are in the user defined objective function.
Taken together, JuPOETs has demonstrated improved flexibility, and performance over POETs in parameter identification and ensemble generation for multiple objectives.
JuPOETs has the potential for widespread use due to the flexibility of the implementation, and the high level syntax and distribution tools native to Julia.

>>>>

\begin{abstract} % abstract
Ensemble modeling is a well established approach for obtaining robust predictions and for simulating course grained population behavior in deterministic mathematical models.
In this study, we present a multiobjective based technique to estimate model ensembles, the Pareto Optimal Ensemble Technique in the Julia programming language (JuPOETs).
JuPOETs integrates simulated annealing with Pareto optimality to estimate parameter ensembles on or near the optimal tradeoff surface between competing training objectives.
We demonstrated JuPOETs on a suite of multiobjective problems, including test functions with parameter bounds and system constraints as well as for the identification of a proof-of-concept biochemical model with four conflicting training objectives.
JuPOETs identified optimal or near optimal solutions approximately six-fold faster than a corresponding implementation in Octave for the suite of test functions.
For the proof-of-concept biochemical model, JuPOETs produced an ensemble of parameters that gave both the mean of the training data for conflicting data sets,
while simultaneously estimating parameter sets that performed well on each of the individual objective functions.
JuPOETs can be adapted to solve many problem types, including mixed binary and continuous variable types, bilevel optimization problems
and constrained problems without altering the base algorithm.
JuPOETs can be installed using the Julia package manager from the JuPOETs GitHub repository at https://github.com/varnerlab/POETs.jl.
\end{abstract}

@@

\section*{Conclusions}
JuPOETs is a significant advance over the previous POETs implementation.
It offers improved performance and is highly adaptable to different problem types.
We demonstrated JuPOETs on a suite of test problems, and a proof-of-concept biochemical model.
However, there are several areas that could be explored further to improve JuPOETs.
First, JuPOETs should be compared with other multiobjective evolutionary algorithms (MOEAs) to determine its relative performance on test and real world problems.
Many evolutionary approaches e.g., the nondominated sorting genetic algorithm (NSGA) family of algorithms, have been adapted to solve multiobjective optimization problems \cite{DEB2002,HUBAND2006}. It is unclear if JuPOETs will perform as well as these other approaches;
one potential advantage that JuPOETs may have is the local refinement step which temporarily reduces the problem to a single objective formulation.
Previously, this hybrid approach led to better convergence on a proof-of-concept signal transduction model \cite{Song:2010aa}.
For many real world parameter estimation problems, the bulk of the execution time is spent evaluating the objective functions.
One strategy to improve performance could be to optimize surrogates \cite{SURROGATES}, while another would be parallel execution of the objective functions.
Currently, JuPOETs serially evaluates the objective function vector.
However, parallel evaluation of the objective functions could be easily implemented using a variety of techniques without changing the main run loop of JuPOETs.
Because of the flexible function pointer architecture of JuPOETs, the only changes required are in the user defined objective function.
Taken together, JuPOETs has demonstrated improved flexibility, and performance over POETs in parameter identification and ensemble generation for multiple objectives.
JuPOETs has the potential for widespread use due to the flexibility of the implementation, and the high level syntax and distribution tools native to Julia.
